Visualizing the Geography of Genetic Variants

### 2016-06-30 paper.Rmd
added more text, nearly draft finished figures and started fig captions

# Abstract {-}
One of the key characteristics of any genetic variant, beyond its potential phenotypic effects or its frequency, is its geographic distribution. The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations spread within, and in turn can help one learn about historical patterns of migration, genetic drift, and natural selection. Collectively the geographic distribution of genetic variants can help to explain how populations have been related through time (e.g. levels of gene flow and divergence). For these reasons, visual inspection of geographic maps for genetic variants is common practice in genetic studies. Here we develop an interactive visualization tool for illuminating the geographic distribution of genetic variants. Through an efficient RESTful API and dynamic front-end the Geography of Genetic Variants (GGV) browser rapidly provides maps of allele frequencies in populations distributed across the globe.

# Introduction {-}
Genetics researchers often face the following problem: the researcher has identified one or more genetic variants of interest (e.g. from a genome-wide association, eQTL, or pharmacogenomic study) and would like to know the geographic distribution of the variant. For example, the researcher may hope to address: 1) implications for genomic medicine (e.g. is a risk allele geographically localized to a certain patient population?) @XXX; 2) design of follow-up studies (e.g. what population should be studied to observe variant carriers?) @XXX; or 3) the biology of the variant in question (e.g. does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?) @novembre2009spatial.  A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.

@@

Our work is inspired by past tools such as the ALFRED database @rajeevan2011alfred and the maps available on the HGDP Selection browser @pickrell2009signals.  One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts with a sbustantial learning curve.  The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g @XXX, books e.g. @dudley2013exploring, and have been ported and made avialable on the UCSC Genome Browser (available under the XX track of the browser).

@@

In addition, novel approaches for data visualization have become more widely available.  In particular, web-based visualization tools, such as Data Driven Documents (d3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of share-ability @bostock2011d3.

@@

# Fundamental Approach {-}
The Geography of Genetic Variants browser (GGV) uses the SVG and mapping utilities of d3.js to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.

In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project, Human Genome Diversity project and POPRES project, we have developed an SQL database and RESTful API for querying allele frequencies by chromosome and position or reference SNP identifier [Figure 1].

```{r figure1, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Figure caption'}
img <- readPNG("figures/png/figure1.png")
grid.raster(img)
```

Users can query single variants by chromosome and position identifiers, by rsids @sherry2001dbsnp, or users can simply choose a random variant from within a dataset.  While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset. For instance, in data with deep population subdivision, it is obvious in the consistent patterns of differentiation observed across markers. We also expect this will be useful in teaching contexts – as it provides a highly visual way for learners to understand human genetic variation.

After a query, the GGV displays the frequency of a variant in a given population as a pie chart where each slice represents minor and major allele frequencies. Pie charts are displayed as points positioned at the population's region of origin where the projection/map-view is chosen based off of the geographic proximity of populations present in a given dataset [Figure 2].

```{r figure2, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Figure caption'}
img <- readPNG("figures/png/figure2.png")
grid.raster(img)
```

@@

# Representing rare variants in frequency data {-}
An additional challenge is that allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset. This has not been an pervasive problem until recently, as most population genetic samples were based on genotype arrays biased towards variants that are common in human populations (5-50% in minor allele frequency). With the advent of next generation sequencing technologies and large samples of thousands of individuals, it is common for datasets to contain rare variants @10002015global, @nelson2012abundance, @tennessen2012evolution.

@@

# Underlying Frequency API {-}
[Joe - worth saying more about the API structure and giving example calls and structure of JSON?]

@@

# Acknowledgements {-}
Support for this work was provided by the NIH-BD2K initiative (1U01 CA198933-0).  The authors would also like to thank XX for supportive conversations.

# Acknowledgements {-}

>>>>

# Abstract {-}
One of the key characteristics of any genetic variant, beyond its potential phenotypic effects or its frequency, is its geographic distribution. The geographic distribution of a genetic variant can shed light on where the variant first arose, in what populations it spread within, and in turn can help one learn about historical patterns of migration, genetic drift, and natural selection. Collectively the geographic distribution of genetic variants can help to explain how populations have been related through time (e.g. levels of gene flow and divergence). For these reasons, visual inspection of geographic maps for genetic variants is common practice in genetic studies. Here we develop an interactive visualization tool for illuminating the geographic distribution of genetic variants. Through an efficient REST API and dynamic front-end the Geography of Genetic Variants (GGV) browser rapidly provides maps of allele frequencies in populations distributed across the globe.

# Introduction {-}
Genetics researchers often face the following problem: the researcher has identified one or many genetic variants of interest (e.g. from a genome-wide association, eQTL, or pharmacogenomic study) and would like to know the geographic distribution of the variant. For example, the researcher may hope to address: 1) implications for genomic medicine (e.g. is a risk allele geographically localized to a certain patient population?); 2) design of follow-up studies (e.g. what population should be studied to observe variant carriers?); or 3) the biology of the variant in question (e.g. does the variant correlate with a known environmental factor in a manner suggestive of some geographically localized selection pressure?)  @rosenberg2010genome, @novembre2009spatial, @coop2010using.  A simple geographic map of the distribution of a genetic variant can be incredibly insightful for these questions.

The geographic distribution of rare genetic variants (minor allele frequency $\leq .05$) can be particularly informative in determining an individual's ancestry. Rare variants tend to be localized in small geographic regions because they have arisen recently in human history and have had little time to spread across populations through migration @nelson2012abundance, @mathieson2014demography. Thus by surveying the collection of rare variants that an individual carries provides information about the population origins of that individual's recent ancestors. One of us (J.N.) recently took advantage of the geographic clustering of an individual's rare variants in helping to reaffirm the historically highly contested Native American ancestry of the Kennewick Man, an ancient individual from Washington State @KennewickManReport.

@@

Our work is inspired by past tools such as the ALFRED database @rajeevan2011alfred and the maps available on the HGDP Selection browser @pickrell2009signals.  One of us (J.N.) developed the scripts for the HGDP Selection Browser maps using The Generic Mapping Tools (GMT) @wessel2013generic, a powerful system of geographic plotting scripts with a sbustantial learning curve.  The plots from the HGDP Selection Browser have proved useful, and have appeared in research articles e.g @pickrell2009signals, @coop2009role, books e.g. @dudley2013exploring, and have been ported and made avialable on the UCSC Genome Browser (available under the HGDP Allele Freq track of the browser).

@@

In addition, novel approaches for data visualization have become more widely available.  In particular, web-based visualization tools, such as Data Driven Documents (D3.js), offer powerful approaches to realize these aims through useful methods for interactivity, advantages of software development in modern web-browsers, a large open-source development community, and ease of share-ability @bostock2011d3.

@@

# Fundamental Approach {-}

The Geography of Genetic Variants browser (GGV) uses the SVG and mapping utilities of D3.js to generate interactive frequency maps, allowing for quick and dynamic displays of the geographic distribution of a genetic variant.

```{r figure1, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Schematic diagram of the Geography of Genetic Variants (GGV) web application. We compute minor allele frequencies from publicly available population genetic datasets from the 1000 Genomes Project, Human Genom Diversity Project (HGDP) and Population Reference Sample (POPRES). These allele freqeuncies are stored in a SQL database along side with geographic coordinates associated with each population. The front-end visualization was created using d3 and accesses allele frequency and population meta-data from a REST API implemented in the python library flask.'}
img <- readPNG("figures/png/figure1.png")
grid.raster(img)
```

In order to allow for easy access to large commonly used public genomic datasets, such as the 1000 Genomes project, Human Genome Diversity project and POPRES project, we have developed an SQL database and REST API for querying allele frequencies by chromosome and position or reference SNP identifier [Figure 1].

Users can query single variants by chromosome and position identifiers, by rsid @sherry2001dbsnp, or users can simply choose a random variant from within a dataset.  While many applications require inspection of the distribution of a specific variant, from our experience, it can be very helpful to quickly view the geographic distribution of several randomly chosen variants to quickly gain a sense of structure in a dataset. For instance, in data with deep population subdivision, it is obvious in the consistent patterns of differentiation observed across markers. We also expect this will be useful in teaching contexts – as it provides a highly visual way for learners to understand human genetic variation.

```{r figure2, echo=FALSE, message=FALSE, fig.height=5, fig.width=5, fig.align='center', fig.cap='Example map from the Geography of Genetic Variants web application. Each pie chart represents a population with the blue slice of the pie displaying the freqeuncy of the global minor allele and the yellow slice of the pie displaying the frequency of the global major allele in each population.'}
img <- readPNG("figures/png/figure2.png")
grid.raster(img)
```

After a query, the GGV displays the frequency of a variant in a given population as a pie chart where each slice represents minor and major allele frequencies. Pie charts are displayed as points positioned at the population's region of origin where the projection / map-view is chosen based off of the geographic proximity of populations present in a given dataset [Figure 2].

@@

# Representing rare variants in frequency data {-}
An additional challenge is that allele frequencies between variants often differ greatly, sometimes by orders of magnitude in a single dataset. This has not been an pervasive problem until recently, as most population genetic samples were based on genotype arrays, biased towards variants that are common in human populations (5-50% in minor allele frequency). With the advent of next generation sequencing technologies and large samples of thousands of individuals, it is common for datasets to contain rare variants @10002015global, @nelson2012abundance, @tennessen2012evolution.

@@

# Underlying Frequency API {-}
To interface population minor allele frequency data stored in the SQL database with the front-end D3.js visualization we use a REST API impelemented in the python library Flask-RESTful @citeXXX. GET requests are submitted to the database via HTTP, returning json formatted allele frequency data and the meta-data associated with each population and genetic variant e.g. latitude, longitude, population label, sample size, and frequency scale. Genetic variants can be queried by chromosome position, rsid, or randomly. Example HTTP requests and json response can be seen below:

## Query API by rsid {-}
```
http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&rsID=rs1834640
[
  {
    "alleles": ["A", "G"], 
    "pos": ["-15.310139", "13.443182"], 
    "pop": "GWD", 
    "nobs": "226", 
    "xobs": "17", 
    "freqscale": 1, 
    "freq": [0.0752212389381, 0.9247787610619], 
    "chrom_pos": "15:48392165", 
    "rawfreq": 0.0752212389381
  },
  ...
]
```

## Query API by chromosome position {-}

```
http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&chr=14&pos=37690093
[
  {
    "alleles": ["G", "A"], 
    "pos": ["-15.310139", "13.443182"], 
    "pop": "GWD", 
    "nobs": "226", 
    "xobs": "0", 
    "freqscale": 0.01, 
    "freq": [0.0, 1.0], 
    "chrom_pos": "14:37690093", 
    "rawfreq": 0.0
  }
  ...
]
```

## Randomly query API {-}

```
http://popgen.uchicago.edu/ggv_api/freq_table?data="1000genomes_phase3_table"&random_snp=True
[
  {
    "alleles": ["T", "C"], 
    "pos": ["-15.310139", "13.443182"], 
    "pop": "GWD", 
    "nobs": "226", 
    "xobs": "0", 
    "freqscale": 0.01, 
    "freq": [0.0, 1.0], 
    "chrom_pos": "5:42452893", 
    "rawfreq": 0.0
  }, 
  ...
]
```

@@

# Acknowledgements {-}
Support for this work was provided by the NIH-BD2K initiative (1U01 CA198933-0).  The authors would also like to thank members of the Novembre Lab for supportive conversations.
